---
title: "categoryCompare2 Executable Scripts"
author: "Robert M Flight"
date: "`r Sys.time()`"
---

<!--
% \VignetteEngine{knitr}
% \VignetteIndexEntry{categoryCompare2: Executable Scripts}
% \VignetteDepends{categoryCompare2, GO.db, KEGG.db, affy, hgu95av2.db, genefilter, estrogen, limma, DiagrammeR, visNetwork}
% \VignetteKeyword{categoryCompare, meta-analysis, gene ontology, kegg}
% \VignettePackage{categoryCompare2}
-->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
root_loc <- rprojroot::find_root("DESCRIPTION")

tmp_loc <- tempdir()
Sys.setenv(file_loc = root_loc)
Sys.setenv(exec_loc = file.path(root_loc, "inst/executables"))
Sys.setenv(test_loc = file.path(root_loc, "inst/extdata/test_data"))
Sys.setenv(results_loc = tmp_loc)

Sys.chmod(dir(file.path(root_loc, "inst", "executables"), pattern = "*.R", full.names = TRUE), "0750")
```

## Scripts??

This version of `categoryCompare2` now includes a set of executable scripts that
can be used to perform an analysis from the command line. This document attempts
to document what these scripts do. The scripts are found in the folder `exec`
from the package directory.

You can run the `install_executables` function to move the scripts somewhere more
useful, or `executable_path` to provide the locations of the scripts so that you
can make aliases and change their permissions yourself so that
they are accessible from the command line directly.

## Workflow

### Feature Lists

`categoryCompare2` assumes you have one or more feature lists (normally genes) that you want to
run enrichment on and compare the enrichments from the feature lists. For enrichment,
one also needs the full list or **universe** of features that were measured in
the experiment. 

The easiest way to generate the input file in the expected format is to provide
newline separated feature lists, each one in a separate file, where the file name
indicates what condition the features came from.

In this example, we have the gene **symbols** for the differential genes at two
different timepoints, **10** and **48** hours. These were calculated using the
`estrogen` dataset from Bioconductor.


```{sh show_symbols_10}
head $test_loc/10_symbol.txt
```

```{sh show_symbols_48}
head $test_loc/48_symbol.txt
```

There is also a file with the full set of features measured on the Affymetrix
chip for this data set.

To generate the feature file, we need to combine these individual files into
a single JSON file. This uses `feature_files_2_json.R`

```{sh show_help}
$exec_loc/feature_files_2_json.R --help
```

Let's run it!

```{sh run_features}
$exec_loc/feature_files_2_json.R --json="$results_loc/features.json" \
  --file1="$test_loc/10_symbol.txt" \
  --file2="$test_loc/48_symbol.txt" \
  --universe="$test_loc/universe_symbol.txt"
```

Now we can see that the `features.json` file has the three sets of gene symbols
in it.

```{sh show_features_json}
head $results_loc/features.json
```

### Annotations

```{sh annotation_help}
$exec_loc/create_annotations.R --help
```

#### Organism DB

Often times you will want to generate annotations from an included organism
database that generated and maintained by Bioconductor. These are *normally*
the `org.*.eg.db`, also known as organism entrez gene databases. There are
other databases, including the various measurement platform databases.

See `?get_db_annotation` for all the various feature types and annotation types
available.

We will create an annotation object using the `cellular component` sub-ontology
of the Gene Ontology mapped to gene *symbols*.

```{sh create_annotation}
$exec_loc/create_annotations.R --orgdb="org.Hs.eg.db" \
  --feature-type="SYMBOL" \
  --annotation-type="CC" \
  --json="$results_loc/annotations.json"
```

#### User Provided Annotations

If you have annotations you want to use, they should be in a JSON file where
each JSON object is the feature, with the annotations that it has.

An example would be:

```{sh show_custom_annotation}
head $test_loc/create_annotations_example_input.json
```

Using this file, you would run:

```sh
$exec_loc/create_annotations.R --input="create_annotations_example_input.json" \
  --feature-type="SYMBOL" \
  --annotation-type="CC" \
  --json="$results_loc/annotations.json"
```

In this case the `--feature-type` and `--annotation-type` should reflect what
the original sources that were used to generate these annotations.

### Comparisons!

Finally, with the features and annotations in hand, we can do the enrichments
and comparisons between them using `run_enrichments.R`.

```{sh run_enrichments_help}
$exec_loc/run_enrichments.R --help
```

So, with our generated files, we would now do:

```{sh run_enrichments}
$exec_loc/categoryCompare2.R --features="$results_loc/features.json" \
  --annotations="$results_loc/annotations.json" \
  --p-cutoff=0.001 \
  --output-directory="$results_loc" 
```

Which provides the output in `r file.path(tmp_loc, "full_table.txt")`. Here is a preview
of the results:

```{r show_results, echo=FALSE, results='asis'}
in_results <- read.table(file.path(tmp_loc, "full_table.txt"), sep = "\t", header = TRUE)
knitr::kable(head(in_results), digits = 2)
```
