---
title: "categoryCompare2: Checking Executables"
author: "Robert M Flight"
date: "`r Sys.time()`"
---

<!--
% \VignetteEngine{knitr}
% \VignetteIndexEntry{categoryCompare2: Checking Executables}
% \VignetteDepends{categoryCompare2, GO.db, KEGG.db, affy, hgu95av2.db, genefilter, estrogen, limma, DiagrammeR, visNetwork}
% \VignetteKeyword{categoryCompare, meta-analysis, gene ontology, kegg}
% \VignettePackage{categoryCompare2}
-->


```{r setup}
knitr::opts_chunk$set(echo = TRUE)
root_loc <- rprojroot::find_root("DESCRIPTION")

tmp_loc <- tempdir()
Sys.setenv(exec_loc = system.file("exec", package = "categoryCompare2"))
Sys.setenv(test_loc = system.file("extdata", "test_data", package = "categoryCompare2"))
Sys.setenv(results_loc = tmp_loc)

Sys.chmod(dir(file.path(root_loc, "inst", "executables"), pattern = "*.R", full.names = TRUE), "0750")
library(categoryCompare2)
library(tools)
```

## Purpose

Verify that the executables give the same results as running `categoryCompare` itself.

## R Version

We will use our R programming to read in the data and generate the annotations.

```{r read_in_data}
get_feature_lists <- function(file_list){
  file_not_universe <- unlist(file_list[!(names(file_list) %in% "universe")])
  
  condition_names <- basename(file_not_universe)
  condition_names <- gsub(paste0(".", file_ext(condition_names[1])), "", condition_names)
  
  file_data <- lapply(file_not_universe, function(x){
    readLines(x)
  })
  names(file_data) <- condition_names
  
  if (is.null(file_list$universe)) {
    file_data$universe <- unique(unlist(file_data))
  } else {
    file_data$universe <- readLines(file_list$universe)
  }
  
  file_data
}

file_list <- list(file1 = file.path(test_loc, "10_symbol.txt"), universe = file.path(test_loc, "universe_symbol.txt"))
feature_list <- get_feature_lists(file_list)
feature_universe <- feature_list$universe
feature_list$universe <- NULL
```

```{r generate_annotations}
annotation_obj <- get_db_annotation("org.Hs.eg.db", feature_type = "SYMBOL", annotation_type = "CC")
```

```{r enrichments_R}
gene_enrichments <- lapply(feature_list, function(in_genes){
  hypergeometric_feature_enrichment(
    new("hypergeom_features", significant = in_genes,
        universe = feature_universe, annotation = annotation_obj),
    p_adjust = "BH"
  )
})

combined_enrichments <- combine_enrichments(gene_enrichments)

p_cutoff_column <- "padjust"
p_cutoff_value <- 0.01
p_cutoff_direction <- "<="

count_cutoff_column <- "counts"
count_cutoff_value <- 2
count_cutoff_direction <- ">="

count_call_info <- list(fun = count_cutoff_direction, var_1 = count_cutoff_column, var_2 = count_cutoff_value)
p_call_info <- list(fun = p_cutoff_direction, var_1 = p_cutoff_column, var_2 = p_cutoff_value)

significant_calls <- list(counts = count_call_info, pvalues = p_call_info)

combined_significant <- combined_significant_calls(combined_enrichments, significant_calls)

results_table <- generate_table(combined_significant)
```

## Executable Version

```{sh}
$exec_loc/feature_files_2_json.R --json="$results_loc/features.json" \
  --file1="$test_loc/10_symbol.txt" \
  --universe="$test_loc/universe_symbol.txt"
```

```{sh}
$exec_loc/create_annotations.R --orgdb="org.Hs.eg.db" \
  --feature-type="SYMBOL" \
  --annotation-type="CC" \
  --json="$results_loc/annotations.json"
```

```{sh}
$exec_loc/run_enrichment.R --features="$results_loc/features.json" \
  --output-file="$results_loc/cc2_results.txt" \
  --text-only="TRUE" \
  --annotations="$results_loc/annotations.json"
```

```{sh}
$exec_loc/filter_and_group.R --enrichment-result="$results_loc/cc2_results.txt" \
  --table-file="$results_loc/cc2_results_grouped.txt"
```

## Comparison

```{r read_exec_results}
exec_results <- read.table(file.path(tmp_loc, "full_table.txt"), sep = "\t", header = TRUE, stringsAsFactors = FALSE)
```

```{r join_results}
both_results <- dplyr::full_join(results_table, exec_results)

p_diff <- data.frame(diff = both_results$`10_symbol.p` - both_results$X10_symbol.p)
max(p_diff$diff)
```

```{r visualize_diffs}
library(ggplot2)
sum(is.na(both_results$`10_symbol.p`))
sum(is.na(both_results$X10_symbol.p))
ggplot(p_diff, aes(x = -1*log10(diff))) + geom_histogram(bins = 100)
```

OK, so where there are both GO terms, the differences are on the order of machine
precision, but there are `r sum(is.na(both_results$X10_symbol.p))` GO terms missing
from the executable case. That is not good!

## Missing GO terms

Lets read in the annotation object and see what GO terms are present there compared
to the one we generated.

```{r read_annotations}
json_annotations <- json_2_annotation(file.path(tmp_loc, "annotations.json"))
```


```{r compare_annotations}
all.equal(json_annotations, annotation_obj)
```

Nope, supposedly have the exact same set of annotations.

## Different Genes Measured

```{r load_json_genes}
json_genes <- jsonlite::fromJSON(file.path(tmp_loc, "features.json"))

setdiff(json_genes$`10_symbol`, feature_list$`10_symbol`)
length(json_genes$`10_symbol`)
length(feature_list$`10_symbol`)

setdiff(json_genes$universe, feature_universe)
length(json_genes$universe)
length(feature_universe)
```

